---
title: "WhatsApp Conversational Analysis"
subtitle: "Thread initiation and first respondents"
author: "Shubhankar Tiwari"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  rmdformats::readthedown:
    highlight: kate
    #self_contained: false
    #code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE,message = FALSE,tidy=TRUE,cache=TRUE)
```

## Shiny App

Please check out the [__SHINY APP__](https://shubhstiws.shinyapps.io/Whatsapp_UMN_Shiny/) if you want to skip over the analysis

## Exporting Data from WhatsApp

I used the export functionality from whatsapp to get the text data from my iPhone to my email account. As I was not analyzing media and links I chose to exclude them to decrease the size of the data set. I copied the text data into a csv files using excel and cleared of white spaces and corrected formatting to be able to load it into R.

It was only possible because this was a small data set which could be easily handles by Excel. Most WhatsApp data sets are humongous and require R/ python for processing.

Some of the common functions I used were:

1. Load data using the text connector

2. Parse date and time, convert to military time and extracts hours, minutes and seconds

3. Logical checks to confirm if parsing was correct

4. Format names with special characters and wrong spellings

## Loading libraries

```{r}
suppressPackageStartupMessages({
library(readr)
library(zoo)
library(dplyr)
library(tidyr)
library(chorddiag)
library(htmlwidgets)
library(DT)
})
```


## Loading file

```{r}
df = read.csv('chat_raw.csv',sep=',')
df$date_time = paste(df$date,df$military_time)
df$date_time <- strptime(df$date_time,format="%m-%d-%y %H:%M:%S")
datatable(head(df))

```

## Data Processing

### Calculating time to respond

We need to account for the replies to the person who initiated the thread. This can be calculated by a lag 1 time. The lag1 time is the time stamp of the next chat message in the data set

```{r}
reply = diff(df$date_time,lag=1)
reply[4416] = 0

df = data.frame(cbind(df,reply))
#head(df)
#summary(df)

```


### Assumptions

Once we have the lag we now know the consecutive time diff between messages. I have used a simple rule that __if after posting a message no body responds for 30 minutes (1800 seconds), I believe that people have moved on and any message after 30 minutes is a new thread in itself__

This is a strong assumptions and may bias the analysis as:

1. People may be posting announcements (may or may not require any response)

2. Few people do not instantaneously reply to messages. The respond only during certain periods of the day like morning and evening

But to simplify and start the analysis I felt this __naive assumption__ to be good enough.

```{r}
df$thread = ifelse(df$reply < 1800,1,0)
```

### Marking thread start points

```{r}
##Final df
#df$thread_start
#df$first_respondedt
#df$count

df$tp = lag(df$thread)
#View(df)

thread_start = ifelse(df$tp==0,levels(df$name),NA)
thread_start = as.data.frame(table(thread_start))

```

### Assigning groups to each thread and the people in it

All the messages between two consecutive `0` have been marked as a part of a thread and we get a total `881` unique threads

```{r}
t = df$tp
t[is.na(t)] <- 0
counter = 0
g=c()
g[1] = 999

for (i in 1:length(t)){
  temp = ifelse(t[i]==0, 1,0)
  counter = counter+temp
  g[i]=counter  
}

df = cbind(df,g)
#View(df)
length(unique(df$g))
```

### Processing data for visualization

The format for the visualization is a matrix which has rows as the name of the person who initiated the thread and the columns as the first respondent

#### No of people involved threads

```{r}
df$date_time = as.Date(df$date_time)
# average people involved in a thread
datatable(df %>% group_by(g) %>% summarise(people_involved = n_distinct(name)) %>% ungroup())
```

#### People asking the real questions - no responses

```{r}
# People asking the real questions - no responses
ignored = df %>% distinct(name,g) %>% group_by(g) %>% filter (n_distinct(name) == 1) %>% group_by(name) %>% summarise(no_response = n()) %>% arrange(desc(no_response))

datatable(ignored)
```

#### Data for the chord diagram

```{r}
temp = df %>% distinct(name,g)
#df %>% distinct(name,g) %>% group_by(g) %>% filter(row_number()==1 | row_number()==n())

# thread with thread initiator and first respondent
chartData = df %>% distinct(name,g) %>% group_by(g) %>% filter (n_distinct(name) > 1) %>% filter(row_number()==1 | row_number()==2) %>% select(g,name) %>% group_by(g) %>% mutate(initiate = row_number()) %>% spread(initiate,name) %>% select(thread = g, thread_start = `1`,first_respondent=`2`) %>% group_by(thread_start,first_respondent) %>% summarise(count=n()) %>% spread(first_respondent,count,fill = 0)
```

#### Converting data to matrix format and visualizing

```{r}

temp = chartData

s = temp$thread_start
e = colnames(temp)[2:ncol(temp)]
srm = as.matrix(temp[, -1])
dimnames(srm) = list(Thread_Start = as.vector(s),First_Respondent = as.vector(e))
#mycols = c("#F44336", "#3F51B5", "#4CAF50", "#FFC107")
plot = chorddiag(srm, showGroupnames = TRUE, type = "bipartite", showTicks = FALSE, 
    groupnameFontsize = 10, categorynameFontsize = 16,height = 1000)


```

## Visualization

```{r}
plot
```

### Insights

1. As evident the plot is a mess. There is too much data to understand any pattern. At first I thought of building an animation which could project the increase in density as time progress but even that is hard to understand and draw anything meaningful. Hence I decide to build an interactive  shiny app which could be used to modify the chart for a given time frame. 

2. Please check out the [SHINY APP](https://shubhstiws.shinyapps.io/Whatsapp_UMN_Shiny/) and play around with the dates

3. You can also hover on the chords to get the number of threads to which a particular person has acted as the first respondent

## Next Steps

1. Logical next step would be analyse the behavior of the people responding the most/ initiating the most threads to find how are they different other users (who are second/ third respondents)

2. We can also do topic modelling and sentiment analysis to see what type of threads gets the most responses or get respondent the fastest

3. The animation can be improved and filters like country, Student name etc can be added

## Source

[Github](https://github.com/shubhstiws/WhatsApp_UMN)

